Fast de novo transpoible element annotation

### 2015-07-24 manuscript/BMC/bmc_article.tex

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
\section({Introduction}
Here we introduce RAIDER (Rapid Ab Initio Detection of Elementary
Repeat), designed to enable the quick annoation of transposible
elements and repetitive DNA without requiring a pre-compiled template
library.  Transposible elements, or TEs, are present in almost every
higher order genome, covering as much as 45\% of the human genome and
90\% of the maize genome [CHECK THIS] [CITE].  Given their prevelence,
identiication of TEs is important not only for those who study them, but for
anyone trying to understand a geome.  If not filtered, TEs can trigger
hug numbers of false positives in the course of applying automated
gene finding techniques, as well as blow up runtime.  [CITE: Jiang]
So identifying and masking them in a must for any genome being
studied.

@@

Another line of developement, first proposed by Zheng and Lonardi,
involves the use of {\it elementary repeats} [CITE].  Similar in concept to
the motifes identified by RepeatGluer [CITE], an elementary repeat is
foramlly defined sequence pattern that could be considered a buiding
block (or motif) of a transposable element.  Zheng and Lonardi defined
the concept and proposed an idenficiation algorith, but that algorithm
suffered from a runtime that was quadratic in the size of the genome.
Two other works, He [CITE] and Huo {\it et al.} employed variations of
suffix trees to get linear identification algorithms, but lost
sensitivity in the resence of genomic substitions -- making them less
useful on real data.

Here we describe RAIDER (Rapid Ab Initio Detection of Elementary
Repeats),

>>>>

\newtheorem{definition}{Definition}
\newtheorem{observation}{Observation}
\newtheorem{lemma}{Lemma}

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
\section*{Background}
Here we introduce phRAIDER (Rapid Ab Initio Detection of Elementary
Repeat), designed to enable the quick annoation of transposible
elements and repetitive DNA without requiring a pre-compiled template
library.  Transposible elements, or TEs, are present in almost every
higher order genome, covering as much as 45\% of the human genome and
90\% of the maize genome [CHECK THIS] [CITE].  Given their prevelence,
identiication of TEs is important not only for those who study them, but for
anyone trying to understand a geome.  If not filtered, TEs can trigger
hug numbers of false positives in the course of applying automated
gene finding techniques, as well as blow up runtime.  [CITE: Jiang]
So identifying and masking them in a must for any genome being
studied.

@@

\subsection*{Elementary Repeats}
Another line of developement, first proposed by Zheng and Lonardi,
involves the use of {\it elementary repeats} [CITE].  
Zheng and Lonardi formulated their defintion with the explicit intention of factoring minimum length and
frequency into the definition: to be an elementary repeat a sequence
must have some minimum length $l$ and occur at least $f$ times.  More
importantly, it must be minimal (that is, no section of an elementary
repeat can itself be an elementary repeat), and it must have appear
independently of any other sequence meeting definition of an
elementary repeat.  [CITE]  Formally:
\begin{definition}
Give a length criteria $l$ and a frequence criteria $f$, a sequence
$r$ is an elementary repeat in a genome if:
\begin{enumerate}
\label{ZLDef}
\item $r$ is of length at least $l$.
\item There are at least $f$ copies of $r$ in the genome.
\item There is no proper subsequence $r'$ of $r$ such that $r'$ is of
  length $l$ or grater and $r'$ occurs independently of $r$.  (That
  is, every length $>l$ subsequence of $r'$ occurs only where $r'$
  occurs in the genome.  (The minimality condition.)
\item There is no sequence $r'$ that properly contains $r$ and appears
  around every occurence of $r$ in the genome.
\end{enumerate}
\end{definition}
The third condition insures minimality (that an elementary repeat
cannot contain elementary repeats), while the fourth condition that
each elementary repeat is maximal in length.  Having proposed this
definition, Zheng and Lonardi developed an identification algorithm
that had a runtime quadratic in the genome size -- so not of practical
use on whole-genome analysis.  This was refined to linear time by He [CITE],
and also by Huo {\it et al.} [CITE] based on variations suffix tree
approaches, but because of that use of suffix trees they are limited
in their ability to handle sequence variation.  As transposible
elements naturallly accumulate instance-specific base substitutions
over time, this is a significant limitation.

\subsection*{RAIDER}
It was with the objective of creating a linear time identification
algorithm that could handle variation through use of
PatternHunter-like spaced seeds [CITE].  A rough implementatin was
first presentd in Figuera {\it et al.}, with more details in the
Figuera masters thesis [CITE].  Not itself fully able to handle the
spaced seed component, RAIDER was built along an alternate,
but equivilent definition of elementary repeats based on $l$-mers
(sequences of length $l$).  Specifically, it was observed that
condition (3) of Definition~\ref{ZLDef} could be rewritten as: There
is no $l$-mer contained with $r$ that appears in the genome
independelty of $r$.

From this it follows that:
\begin{enumerate}
\item An $l$-mer cannot belong to two different elementary repeats.
\item Any $l$-mer in the genome that occurs $f$ or more times is
  either an elementary repeat or is contained on one.
\item Any two $l$-ers contained within an elementary repeat must occur
  the same number of times in the genome.
\item Given two sequences in the genome that are {\it maximally
  idenitcal} (that is, cannot be extended in either direction and stil
  be identical), these sequences cannot be contained with in
  elementart repeat.
\end{enumerate}
(For dicussion and proof, see the Figueroa Thesis [CITE]].)

Based on these observations, we discover we can find all elementary
repeats in a single scan of the genome.  Specifically, as we scan from
left to right, we track $l$-mer occurances, and look for repeated
$l$-mer sequences.  When we find the same sequence of $l$-mers
occuring multile times in a row, we can mark it as a tentative family,
then brack it down later if we discover violations of the minimality
condition.  The alogrithm is summerized in Figueroa {\it et al.} [CITE], and
discussed in detail in the Figueroa Thesis [CITE].

Results of the preliminary implementation were promising.  Based on
the Saha {\it et al} analysis putting RepeatMasker as the top {\it de
  novo} search tool, comparisone were aginst that.  RAIDER result
quality was close, but slightly behind -- with runtime
order of magnitudes better.  On human chromosome 22 we saw a
$12\times$ speedup (2344 seconds to 192 seconds), while coverage
actually improved (77\% to 84\%), while on mouse chromsome 19 we saw 
the same speedup with a sinifcant drop in coverage (53\% to 30\%).  On
the full human genome RAIDER ran in $6.3$ hours, while RepeatScout was
unable to complete it run.  For details, see Figueroa {\it et al.} [CITE]

\subsection*{Spaced Seeds}

PatternHunter, first proposed as a very succesful augmentation to
BLAST [CITE PatternHunter,BLAST], is based on the notion of {\it
  spaced seeds}.  In the context of BLAST, instead of looking for a
cntiguous substring match to triger the alignment exention,
PatternHunter uses a more refised match pattern.  For example, the
{\it seed} 1111110111111 would specify that we were looking for an exact
match of two length-six substrings seperated by a single base (which
might or might not match).  In the toy seed $s=110101$, we would say the
strings $AACACA$ and $AAGACA$ because all positions corresponding to a
1 match.  The string $AACACA$ does not math $TACACA$ (with respect to
$s$), as the single miss-match does not correspond to the 1.  Nor does
$AACACA$ match ant substring of $TTAACTCA$.

The incorperations of spaced seeds into BLAST lead to significant
improvements at vertual no cost in runtime [CITE], so it made sence to
incorportat them here.  Specifcially, instead of looking at repeat
elements as a set of identical strings, we could require that they
match with respect to a PatternHunter spaced seed.  The preliminary
version of RAIDER from Figueroa {\it et al.} made some attempt to
incorperate this -- was succesful enough to provide a
proof-of-concept, but a significant amount of work was left to be done
for the developement of phRAIDER (PatternHunter-based RAIDER) -- the
main method in this paper.

\section{Methods}
